---
title: "stroke_prediction"
author: "Mustafain Ali"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on building and deploying a stroke prediction model in R. It contains analysis such as data exploration, summary statistics and building the prediction models. The final report was completed on `r date()`. 

**Data Description:**

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths.

This data set is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relevant information about the patient.


# Task One: Import data and data preprocessing

## Load data and install packages

```{r}
library(tidyverse)
library(janitor)
library(pROC)
library(tidymodels)

#Loading and reading the CSV File
df <- read_csv("healthcare-dataset-stroke-data.csv",  na = c("", "NA", "N/A"))
# There are 6 character type columns and 6 numeric type columns but we convert number like characters to numerics

# Lets clean up the names (like make them lowercases)
df <- clean_names(df)

#Lets remove unnecessary columsn like patient ids
df <- select(df,-id)

#We can make stroke a categorical variable 
df$stroke <- as.factor(df$stroke)

head(df)
glimpse(df)
```


## Describe and explore the data

```{r}
# we can get a basic summary of the numeric columns using summary command
summary(df)

# The % of strokes
table(df$stroke)

#0 means there are no strokes and 1 means there are strokes

prop.table(table(df$stroke))*100
# Terminal shows that only ~5% people experience strokes. 

#Lets look for missing values 
colSums(is.na(df))

#Time to visualize a bit
ggplot(df, aes(x = age))+geom_histogram(bins = 30) + labs(title = "Distribution of Age")

ggplot(df, aes(x = stroke)) + 
  geom_bar() + 
  labs(title = "Stroke Class Counts", x = "Stroke (0=No, 1=Yes)", y = "Count")
```



# Task Two: Build prediction models

```{r}
#Train/Test Split
set.seed(7788)

split_obj <- initial_split(df, prop = 0.8, strata = stroke)

train <- training(split_obj)
test <- testing(split_obj)

#lets verify if the rows sum to 5011
nrow(train)
nrow(test)

#lets verify if split was correct in test and train and whether class balance is similar!
prop.table(table(train$stroke))*100
prop.table(table(test$stroke))*100

#imputing missing values with medians from training set
median_bmi <- median(train$bmi, na.rm = TRUE)
train$bmi[is.na(train$bmi)] <- median_bmi
test$bmi[is.na(test$bmi)] <- median_bmi

#Other things to do is to convert the categorical features into factors!
cat_col <- c("gender", "ever_married", "work_type", "residence_type", "smoking_status")

for (col in cat_col) {
  train[[col]] <- as.factor(train[[col]])
  test[[col]]  <- as.factor(test[[col]])
  test[[col]]  <- factor(test[[col]], levels = levels(train[[col]]))
}
# Now that we have the data ready, lets use LOGISTIC REGRESSION as our first predictive model

model_logr = glm(stroke ~., data = train, family = binomial, weights = ifelse(train$stroke == 1, 2, 1))

#we have added weights which penalize missing stroke more! this is to address the class imabalnce... could have used SMOTe but umm.. this one was simpler. Maybe in scenario 2 I will use SMOTE.

summary(model_logr)
```




# Task Three: Evaluate and select prediction models

```{r}
pred_prob <- predict(model_logr, newdata = test, type = "response")

head(pred_prob)

threshold <- 0.10
pred_class <- ifelse(pred_prob >= threshold, 1, 0)

# now that we have teh prediction probabilities, lets geta confusion matrix to check FP and FNs

conf <- table(Predicted = pred_class, Actual = test$stroke)
conf

#Related mterics computation
TP <- conf["1","1"]
FN <- conf["0","1"]
TN <- conf["0","0"]
FP <- conf["1","0"]

accuracy <- (TP+TN)/(TP+TN+FP+FN)
sensitivity <- TP / (TP + FN) 
specificity <- TN / (TN + FP)

accuracy
sensitivity
specificity

roc_obj <- roc(test$stroke, pred_prob)
auc(roc_obj)


plot(roc_obj, main = "ROC Curve - Logistic Regression")
abline(a=0, b=1, lty = 2)
```



# Task Four: Deploy the prediction model

```{r}
# I am going to deploy this with a simpel function that takes one patients info and returns their predicted probability of stroke

predict_stroke_risk <- function(new_patient_row) {
  predict(model_logr, newdata = new_patient_row, type = "response")
}

#Example making up FAKE patient data
new_patient <- tibble(
  gender = factor("Male", levels = levels(train$gender)),
  age = 37,
  hypertension = 1,
  heart_disease = 0,
  ever_married = factor("Yes", levels = levels(train$ever_married)),
  work_type = factor("Private", levels = levels(train$work_type)),
  residence_type = factor("Urban", levels = levels(train$residence_type)),
  avg_glucose_level = 160,
  bmi = 28.5,
  smoking_status = factor("formerly smoked", levels = levels(train$smoking_status)),
)

predict_stroke_risk(new_patient)
```




# Task Five: Findings and Conclusions

- The dataset contains both demographic and clinical features and the data distribution shows that stroke is a rare outcome. This fact suggests that there is a class imabalnce. 
- We did a 80/20 train/test data split based on stratification by stroke.
- I used logistic regression to estimate stroke probability. It was an arbitrary choice. Ideally we should test different models and decide based on evaluation metrics which one should be used. 
- Model performance was evaluated using the confusion matric and ROC-AUC. 
- By changing the parameters of the new patient data, once can get an estinate of teh risk of stroke to that person. 

**RESULTS: interpretation**
- AS we can see, the confusion matrix predicts that 31 cases were corretcly detected, 15 were missed, but there were 250 false positives which is a lot so we might need to do some more thresholds/weight adjustement. That said, 726 non-strokes were correctly identified. 
- A 67% sensitivity means I was able to catch 67% of the real stroke cases which is good from a medical screening perspective but it should be higher!
- The 74% specificity and accuracy are tradeoffs for higher sensitivity due to the weighted schemes used.
- An AUC of 80% reflects that if we randomly choose one stroke and one non-stroke patient, my model will rank the stroke patient higher 80% of the time. Looks solid enough!
- The ROC curve is clearly above the diagonal, no visible breaks. 


